---
title: *Hemigrapsus* exploratory statistics
---


```{r}

# Load libraries
library(vegan)      # For diversity analysis
library(ggplot2)    # For basic plotting
library(tidyr)      # For data manipulation
library(dplyr)      # For data manipulation
library(ggpubr)     # For arranging multiple plots
library(reshape2)   # For reshaping data
```

```{r}

#getwd()

# Read-in the data
data <- read.csv("/Users/cmantegna/Documents/YellowIsland2024/data/hemi_complete.csv")
```

```{r}
  
# Explore data
#head(data)
summary(data)
```

# Visualizations for All Crabs

## Carapace Width Distribution
```{r}

# Distribution of Carapace Width
ggplot(data, aes(x = carapace_width)) +
  geom_histogram(binwidth = 1, fill = "blue", color = "black") +
  labs(title = "Distribution of Carapace Width", x = "Carapace Width (mm)", y = "Frequency")

```

## Primary Color Distribution
```{r}

# Distribution of Primary Colors
ggplot(data, aes(x = primary_color)) +
  geom_bar(fill = "skyblue", color = "black") +
  labs(title = "Distribution of Primary Colors", x = "Primary Color", y = "Count") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

## Carapace Size Across Tidal Zones
```{r}

# Distribution of Carapace Size Across Tidal Zones
ggplot(data, aes(x = carapace_width, fill = tidal_zone)) +
  geom_bar(position = "dodge") +
  labs(title = "Distribution of Carapace Size Across Tidal Zones", x = "Carapace Width (mm)", y = "Count") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

## Primary Colors Across Tidal Zones
```{r}

# Distribution of Primary Colors Across Tidal Zones
ggplot(data, aes(x = primary_color, fill = tidal_zone)) +
  geom_bar(position = "dodge") +
  labs(title = "Distribution of Primary Colors Across Tidal Zones", x = "Primary Color", y = "Count") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

##  Distribution of Leg Patterns
```{r}

ggplot(data, aes(x = leg_pattern)) +
  geom_bar(fill = "pink", color = "black") +
  labs(title = "Distribution of Leg Patterns", x = "Leg Pattern", y = "Count") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

## Distribution of Leg Patterns Across Tidal Zones
```{r}

# Distribution of Leg Patterns Across Tidal Zones
ggplot(data, aes(x = leg_pattern, fill = tidal_zone)) +
  geom_bar(position = "dodge") +
  labs(title = "Distribution of Leg Patterns Across Tidal Zones", x = "Leg Pattern", y = "Count") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

## Leg Hair Presence
```{r}

ggplot(data, aes(x = leg_hair)) +
  geom_bar(fill = "orange", color = "black") +
  labs(title = "Presence of Leg Hair", x = "Leg Hair", y = "Count")
```

## Presence of Mottling
```{r}

ggplot(data, aes(x = mottled)) +
  geom_bar(fill = "red", color = "black") +
  labs(title = "Mottled Carapace", x = "Mottled", y = "Count")
```

## Presence of Spotted Claws
```{r}

ggplot(data, aes(x = spotted_claws)) +
  geom_bar(fill = "yellow", color = "black") +
  labs(title = "Spotted Claws", x = "Spotted Claws", y = "Count")
```

## Counts of Crabs by Sex
```{r}


ggplot(data, aes(x = sex)) +
  geom_bar(fill = "purple", color = "black") +
  labs(title = "Counts of Crabs by Sex", x = "Sex", y = "Count")

```

## Unique color combinations
```{r}

# Creating a new column for color combinations
crab_data <- data %>%
  mutate(color_combination = paste(primary_color, secondary_color, sep = "_"))

# Counting the occurrences of each color combination
color_combination_counts <- crab_data %>%
  count(color_combination, sort = TRUE)

# Plotting the color combinations and their frequencies
ggplot(color_combination_counts, aes(x = reorder(color_combination, n), y = n)) +
  geom_bar(stat = "identity", fill = "purple", color = "black") +
  labs(title = "Color Combinations and Their Frequencies", x = "Color Combination", y = "Frequency") +
  theme(axis.text.x = element_text(angle = 90, hjust = 1))

# Number of unique color combinations
unique_color_combinations <- nrow(color_combination_counts)
unique_color_combinations

# Exporting the unique color combinations to a CSV file
write.csv(color_combination_counts, "unique_color_combinations.csv", row.names = FALSE)

```

# Stats
## Chi-squared
```{r}

# Chi-Square Test for Categorical Variables
chi_square_test <- function(df, col1, col2) {
  tbl <- table(df[[col1]], df[[col2]])
  chisq.test(tbl)
}

# Chi-Square Tests
chi_square_results <- list(
  "Leg Hair vs Tidal Zone" = chi_square_test(data, "leg_hair", "tidal_zone"),
  "Mottled vs Tidal Zone" = chi_square_test(data, "mottled", "tidal_zone"),
  "Spotted Claws vs Tidal Zone" = chi_square_test(data, "spotted_claws", "tidal_zone"),
  "Leg Hair vs Sex" = chi_square_test(data, "leg_hair", "sex"),
  "Mottled vs Sex" = chi_square_test(data, "mottled", "sex"),
  "Spotted Claws vs Sex" = chi_square_test(data, "spotted_claws", "sex"),
  "Spotted Claws vs Leg Hair" = chi_square_test(data, "spotted_claws", "leg_hair")
)

chi_square_results

# Leg hair v tidal zone: X-squared = 8.9439, df = 2, p-value = 0.01143
# Mottled v tidal zone: X-squared = 5.9447, df = 2, p-value = 0.05118
# Spotted claw v tidal zone:X-squared = 1.3445, df = 2, p-value = 0.5106 NOT SIGNIFICANT
# Leg hair v sex: X-squared = 1.2368, df = 2, p-value = 0.5388 NOT SIGNIFICANT
# Mottled v sex: X-squared = 1.643, df = 2, p-value = 0.4398 NOT SIGNIFICANT
# Spotted  claw v sex: X-squared = 3.7074, df = 2, p-value = 0.1567 OT SIGNIFICANT
# Spotted claw v leg hair: X-squared = 4.057, df = 1, p-value = 0.04399 


```

## Kruskal- Wallis + Post hoc
```{r}

# Kruskal-Wallis Post Hoc Test
kruskal_posthoc <- function(df, continuous_var, categorical_var) {
  pairwise.wilcox.test(df[[continuous_var]], df[[categorical_var]], p.adjust.method = "bonferroni")
}

# Kruskal-Wallis Tests and Post Hoc Analysis
kruskal_results <- list(
  "Carapace Width vs Tidal Zone" = list(
    test = kruskal_test(data, "carapace_width", "tidal_zone"),
    posthoc = kruskal_posthoc(data, "carapace_width", "tidal_zone")
  ),
  "Carapace Width vs Sex" = list(
    test = kruskal_test(data, "carapace_width", "sex"),
    posthoc = kruskal_posthoc(data, "carapace_width", "sex")
  ),
  "Carapace Width vs Leg Hair" = list(
    test = kruskal_test(data, "carapace_width", "leg_hair"),
    posthoc = kruskal_posthoc(data, "carapace_width", "leg_hair")
  ),
  "Carapace Width vs Claw Spots" = list(
    test = kruskal_test(data, "carapace_width", "spotted_claws"),
    posthoc = kruskal_posthoc(data, "carapace_width", "spotted_claws")
)
)

kruskal_results

```

# Leg Hair
## Make data frames
```{r}

#Making the df for plotting
#crabs_with_leg_hair <- subset(data, leg_hair == 'Y')
#crabs_without_leg_hair <- subset(data, leg_hair == 'N')

#summary(crabs_with_leg_hair) # Mean carapace width is 6.9 (above total average)
#summary(crabs_without_leg_hair) # Mean carapace width is 5.429 (below total average)

```

## Carapace Width Distribution by Presence of Leg Hair (brown/ dark gold is where the two categories overlap)
```{r}

#Carapace Width
# With Leg Hair: The mean carapace width is 6.9 with a standard deviation of 2.33. The range is from 1.0 to 13.0.
# Without Leg Hair: The mean carapace width is 5.43 with a standard deviation of 2.85. The range is from 1.0 to 16.5. 
# Based on the KW test and post hoc, there is a significant difference in size between crabs with leg hair and crabs without.

ggplot() +
  geom_histogram(data = crabs_with_leg_hair, aes(x = carapace_width, fill = "With Leg Hair"), binwidth = 1, alpha = 0.5) +
  geom_histogram(data = crabs_without_leg_hair, aes(x = carapace_width, fill = "Without Leg Hair"), binwidth = 1, alpha = 0.5) +
  labs(title = "Carapace Width Distribution", x = "Carapace Width", y = "Frequency") +
  scale_fill_manual(name = "Leg Hair", values = c("With Leg Hair" = "darkorchid4", "Without Leg Hair" = "gold2")) +
  theme_minimal()

```

## Tidal Zone Distribution by Presence of Leg Hair
```{r}

# Tidal Zone
# With Leg Hair: Majority are found in LOW tidal zone (61 out of 145).
# Without Leg Hair: Majority are found in MID tidal zone(60 out of 162)
# Based on the chi- squared, the positve relationship between L tide and leg hair is significant. Explain this in your talk.
# Frequency doesn't equal 100 because each zone is out of 102, except low and that is out of 103

tidal_zone_counts <- data %>%
  group_by(tidal_zone, leg_hair) %>%
  tally()

ggplot(tidal_zone_counts, aes(x = tidal_zone, y = n, fill = leg_hair)) +
  geom_bar(stat = "identity", position = "stack") +
  geom_text(aes(label = n), position = position_stack(vjust = 0.5), color = "white") +
  labs(title = "Tidal Zone Distribution", x = "Tidal Zone", y = "Frequency") +
  scale_fill_manual(name = "Leg Hair", values = c("Y" = "darkorchid4", "N" = "gold3"), labels = c("With Leg Hair", "Without Leg Hair")) +
  theme_minimal()


```

## Spotted Claw Distribution by Presence of Leg Hair
```{r}

#Spotted Claws:
# With Leg Hair: Most have NO spotted claws (90 out of 145).
#Without Leg Hair: Most have NO spotted claws (119 out of 162).
# Based on the chi-squared results we have a significant positive relationship

spotted_claws_counts <- data %>%
  group_by(spotted_claws, leg_hair) %>%
  tally()

ggplot(spotted_claws_counts, aes(x = spotted_claws, y = n, fill = leg_hair)) +
  geom_bar(stat = "identity", position = "stack") +
  geom_text(aes(label = n), position = position_stack(vjust = 0.5), color = "white") +
  labs(title = "Spotted Claws Distribution", x = "Spotted Claws", y = "Frequency") +
  scale_fill_manual(name = "Leg Hair", values = c("Y" = "darkorchid4", "N" = "gold3"), labels = c("With Leg Hair", "Without Leg Hair")) +
  theme_minimal()


```
## Sex Distribution by Presence of Leg Hair
```{r}
# Sex Distribution

sex_counts <- data %>%
  group_by(sex, leg_hair) %>%
  tally()

ggplot(sex_counts, aes(x = sex, y = n, fill = leg_hair)) +
  geom_bar(stat = "identity", position = "stack") +
  geom_text(aes(label = n), position = position_stack(vjust = 0.5), color = "white") +
  labs(title = "Sex Distribution", x = "Sex", y = "Frequency") +
  scale_fill_manual(name = "Leg Hair", values = c("Y" = "darkorchid4", "N" = "gold3"), labels = c("With Leg Hair", "Without Leg Hair")) +
  theme_minimal()

```

# Claw Spots
## Carapace Width Distribution by Presence of Spotted Claws
```{r}
# With Spotted Claws: The mean carapace width is 7.79 with a standard deviation of 2.68. The range is from 3.0 to 16.5
# Without Spotted Claws: The mean carapace width is 5.43 with a standard deviation of 2.36. The range is from 1.0 to 12.5. 
# Based on the KW test and post hoc, there is a significant difference in size between crabs with claw spots and crabs without.

ggplot(data, aes(x = carapace_width, fill = spotted_claws)) +
  geom_histogram(binwidth = 1, alpha = 0.5, position = "identity") +
  labs(title = "Carapace Width Distribution Based on Spotted Claws", x = "Carapace Width", y = "Frequency") +
  scale_fill_manual(name = "Spotted Claws", values = c("Y" = "darkorchid4", "N" = "gold3"), labels = c("With Spotted Claws", "Without Spotted Claws")) +
  theme_minimal()

```

## Tidal Zone Distribution by Presence of Claw Spots
```{r}

# Calculate the counts for each category
tidal_zone_counts <- data %>%
  group_by(tidal_zone, spotted_claws) %>%
  tally()

# Plotting the stacked bar chart with counts
ggplot(tidal_zone_counts, aes(x = tidal_zone, y = n, fill = spotted_claws)) +
  geom_bar(stat = "identity", position = "stack") +
  geom_text(aes(label = n), position = position_stack(vjust = 0.5), color = "white") +
  labs(title = "Tidal Zone Distribution by Presence of Claw Spots", x = "Tidal Zone", y = "Frequency") +
  scale_fill_manual(name = "Claw Spots", values = c("Y" = "darkorchid4", "N" = "gold3"), labels = c("With Claw Spots", "Without Claw Spots")) +
  theme_minimal()

```

```{r}

library(dplyr)

# Calculate the counts for each category
both_claw_spots_and_leg_hair <- data %>% filter(spotted_claws == 'Y' & leg_hair == 'Y') %>% nrow()
neither_claw_spots_nor_leg_hair <- data %>% filter(spotted_claws == 'N' & leg_hair == 'N') %>% nrow()
leg_hair_without_claw_spots <- data %>% filter(spotted_claws == 'N' & leg_hair == 'Y') %>% nrow()
claw_spots_without_leg_hair <- data %>% filter(spotted_claws == 'Y' & leg_hair == 'N') %>% nrow()

# Create the sizes object
sizes <- c(both_claw_spots_and_leg_hair, neither_claw_spots_nor_leg_hair, leg_hair_without_claw_spots, claw_spots_without_leg_hair)

# Print the sizes to check the values
print(sizes)

```


## Distribution of Crabs Based on Claw Spots and Leg Hair, Pie Chart
```{r}

library(ggplot2)

# Define the data
sizes <- c(55, 119, 90, 43)  # counts for each category
labels <- c('Both Claw Spots and Leg Hair', 'Neither Claw Spots nor Leg Hair', 'Leg Hair without Claw Spots', 'Claw Spots without Leg Hair')
colors <- c('#8B0000', '#8B4513', '#CDAD00', '#68228B')  # rust, light brown, gold, purple

# Create a data frame
pie_data <- data.frame(
  category = factor(labels, levels = labels),  # Ensures the order is maintained
  count = sizes
)

# Calculate percentages
pie_data$percentage <- round(pie_data$count / sum(pie_data$count) * 100, 1)

# Plot the pie chart
ggplot(pie_data, aes(x = "", y = percentage, fill = category)) +
  geom_bar(width = 1, stat = "identity") +
  coord_polar("y", start = 0) +
  scale_fill_manual(values = setNames(colors, labels)) +  # Match colors to labels
  geom_text(aes(label = paste0(percentage, "%")), position = position_stack(vjust = 0.5), color = "white") +
  theme_void() +
  labs(title = "Distribution of Crabs Based on Claw Spots and Leg Hair")

```

